% =============================================================================
% MenGrowth-Model: Foundation Model Pipeline for Meningioma Growth Forecasting
% BSc Thesis — Technical Report
% =============================================================================
\documentclass[12pt,a4paper]{article}

% --- Encoding and fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% --- Mathematics ---
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{bm}

% --- Tables and figures ---
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{float}

% --- Layout ---
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

% --- References and links ---
\usepackage[colorlinks=true, linkcolor=blue!70!black, citecolor=green!50!black, urlcolor=blue!60!black]{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}

% --- Code listings ---
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny\color{gray},
}

% --- Algorithms ---
\usepackage{algorithm}
\usepackage{algorithmic}

% --- Misc ---
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{siunitx}

% --- Theorem environments ---
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% --- Custom commands ---
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Vol}{\mathbf{z}_{\text{vol}}}
\newcommand{\Loc}{\mathbf{z}_{\text{loc}}}
\newcommand{\Shape}{\mathbf{z}_{\text{shape}}}
\newcommand{\Res}{\mathbf{z}_{\text{res}}}
\newcommand{\dCor}{\operatorname{dCor}}
\newcommand{\dCov}{\operatorname{dCov}}
\newcommand{\dVar}{\operatorname{dVar}}
\newcommand{\KL}{\operatorname{KL}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\LOPO}{\text{LOPO}}
\newcommand{\MOGP}{\text{MOGP}}
\newcommand{\PAMOGP}{\text{PA-MOGP}}
\newcommand{\LME}{\text{LME}}
\newcommand{\HGP}{\text{H-GP}}
\newcommand{\GP}{\mathcal{GP}}
\newcommand{\LoRA}{\text{LoRA}}
\newcommand{\SDP}{\text{SDP}}

% =============================================================================
% Title
% =============================================================================
\title{%
  \textbf{Meningioma Growth Forecasting from Multi-Modal 3D MRI:\\
  A Three-Stage Complexity Ladder}\\[0.8em]
  \large From Volumetric Baselines to Foundation Model Features
}

\author{%
  Mario Pascual-Gonz\'{a}lez\\
  \textit{Department of Computer Science}\\
  \textit{Universidad de Granada}\\
  \texttt{mariopascual@ugr.es}
}

\date{March 2026}

% =============================================================================
% Document
% =============================================================================
\begin{document}

\maketitle

% --- Abstract ---
\begin{abstract}
Meningiomas are the most prevalent primary intracranial tumours, yet predicting
their growth trajectories from longitudinal MRI remains an open challenge due to
heterogeneous growth dynamics and limited longitudinal data. We address the
question: \emph{at what level of model complexity does additional sophistication
stop improving out-of-sample prediction when the longitudinal cohort contains
only $N = 30$ patients?} We organise the investigation as a three-stage
complexity ladder where each stage must earn its place by demonstrably
outperforming the previous one under Leave-One-Patient-Out Cross-Validation
(LOPO-CV). Stage~1 (volumetric baseline) segments tumours via a
domain-adapted BrainSegFounder encoder, extracts whole-tumour volume, and models
scalar log-volume trajectories with Gaussian Process (GP), Linear Mixed-Effects
(LME), and Hierarchical GP models. Stage~2 (latent severity) introduces a
patient-specific severity parameter via a nonlinear mixed-effects model.
Stage~3 (representation learning) tests whether deep foundation model features
add predictive value beyond volume.

Our central finding is that \textbf{parsimony prevails}: the best model is a
simple LME on volumes from a decoder-adapted BrainSegFounder segmentation,
achieving $R^2 = 0.387$ with 90\% calibration coverage. Stage~2 fails to
improve over Stage~1 ($R^2 = -3.54$), largely due to the ordinal time
representation available in our cohort. Stage~3 is not attempted, as the
Stage~2 gate is not passed. We further report that segmentation source is a
stronger determinant of prediction quality than model complexity: the same LME
model yields $R^2 = 0.028$ on manual segmentations versus $R^2 = 0.387$ on
automated segmentations, and that LoRA adaptation achieves only marginal
improvement ($\Delta R^2 = +0.016$) for downstream feature quality while
catastrophically degrading segmentation (Dice $= 0.003$). These findings
establish the first out-of-sample prediction benchmarks for meningioma growth
and demonstrate that the complexity ladder framework effectively identifies the
appropriate level of model sophistication for small longitudinal cohorts.
\end{abstract}

\newpage
\tableofcontents
\newpage

% --- Section 1: Introduction ---
\input{sections/introduction}

% --- Section 2: Background and Related Work ---
\input{sections/background}

% --- Section 3: Data ---
\input{sections/data}

% --- Section 4: Domain Gap Analysis ---
\input{sections/domain_gap}

% --- Section 5: Phase 1 — LoRA Encoder Adaptation (Stage 3 prerequisite) ---
\input{sections/lora_adaptation}

% --- Section 6: Phase 2 — Supervised Disentangled Projection (Stage 3 prerequisite) ---
\input{sections/sdp}

% --- Section 7: Phase 3 — Cohort Encoding and Harmonisation ---
\input{sections/encoding}

% --- Section 8: Growth Prediction — A Three-Stage Complexity Ladder ---
\input{sections/growth_prediction}

% --- Section 9: Evaluation — Variance Decomposition and Quality Assessment ---
\input{sections/evaluation}

% --- Section 10: Conclusion ---
\input{sections/conclusion}

% =============================================================================
% Bibliography
% =============================================================================
\bibliographystyle{unsrt}
\bibliography{references}

\end{document}
