% =============================================================================
% Section 2: Background and Related Work
% =============================================================================
\section{Background and Related Work}
\label{sec:background}

% ---------------------------------------------------------------------------
% 2.1 Meningioma Growth Biology
% ---------------------------------------------------------------------------
\subsection{Meningioma Growth Biology}
\label{sec:bg:growth_biology}

Understanding the biological dynamics of meningioma growth is essential for
selecting appropriate mathematical models. Unlike gliomas, which are
infiltrative parenchymal tumours, meningiomas are extra-axial masses arising
from the arachnoid cap cells of the meninges. They grow by displacing rather
than infiltrating brain tissue, and their dural attachment constrains their
spatial dynamics.

\subsubsection{Growth Patterns and Subgroups}

Engelhardt et al.\ \cite{engelhardt2023gompertz} conducted a comprehensive
analysis of meningioma growth trajectories and found that the Gompertz model
provides the best fit among standard parametric growth curves. They identified
three growth subgroups: (i)~non-growing tumours that remain volumetrically
stable; (ii)~slowly growing tumours with near-linear dynamics over clinical
timescales; and (iii)~rapidly growing tumours displaying clear deceleration
consistent with Gompertz saturation. Approximately 67\% of incidental
meningiomas show measurable growth at 4--5~year follow-up
\cite{hashimoto2012natural}.

Fountain et al.\ \cite{fountain2017growth} compared logistic and Gompertz
models for benign versus atypical meningiomas, confirming that saturating
growth models outperform exponential fits for most patients. The critical
clinical implication is that short-term growth rates do not reliably predict
long-term behaviour: a tumour on the decelerating phase of a Gompertz curve
may appear aggressive early but stabilise later.

\subsubsection{Volumetric versus Diameter Measurements}

Benjamin et al.\ \cite{benjamin2021volumetric} demonstrated that volumetric
assessment is substantially more sensitive than diameter-based measurements for
detecting meningioma growth, with detection rates of 15.2\% versus 5.6\%. This
motivates our use of whole-tumour volume as the primary growth endpoint and
underscores the importance of accurate segmentation.

\subsubsection{Clinical Implications for Modelling}

Three biological facts constrain the modelling framework:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Meningiomas do not migrate.} As extra-axial, dural-attached
    tumours, the spatial centroid is essentially static over time (modulo
    asymmetric growth). Location is therefore a patient-level covariate, not
    a temporal variable.
  \item \textbf{Growth heterogeneity is high.} The three-subgroup structure
    \cite{engelhardt2023gompertz} implies that population-level models must
    accommodate qualitatively different growth regimes, motivating
    mixed-effects or latent variable approaches.
  \item \textbf{Observation density is low.} Typical clinical follow-up
    yields 2--6 scans per patient over months to years, with irregular
    spacing. Models must handle sparse, irregular time series gracefully.
\end{enumerate}

% ---------------------------------------------------------------------------
% 2.2 Gaussian Process Regression for Longitudinal Data
% ---------------------------------------------------------------------------
\subsection{Gaussian Process Regression for Longitudinal Data}
\label{sec:bg:gp}

A Gaussian process (GP) is a distribution over functions such that any finite
collection of function values is jointly Gaussian
\cite{rasmussen2006gaussian}:
\begin{equation}
  f(\mathbf{t}) \sim \GP\bigl(m(\mathbf{t}),\; k(\mathbf{t}, \mathbf{t}')\bigr),
  \label{eq:gp_def}
\end{equation}
where $m(\mathbf{t})$ is the mean function and $k(\mathbf{t}, \mathbf{t}')$
is the covariance (kernel) function. Given $n$ observations
$\mathbf{y} = f(\mathbf{t}) + \boldsymbol{\epsilon}$ with
$\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma_n^2 \mathbf{I})$,
the predictive distribution at test inputs $\mathbf{t}_*$ is:
\begin{align}
  \mathbf{f}_* | \mathbf{t}_*, \mathbf{t}, \mathbf{y}
    &\sim \mathcal{N}(\boldsymbol{\mu}_*, \boldsymbol{\Sigma}_*), \\
  \boldsymbol{\mu}_* &= m(\mathbf{t}_*) + \mathbf{K}_*^\top
    (\mathbf{K} + \sigma_n^2\mathbf{I})^{-1}(\mathbf{y} - m(\mathbf{t})),
    \label{eq:gp_mean} \\
  \boldsymbol{\Sigma}_* &= \mathbf{K}_{**} - \mathbf{K}_*^\top
    (\mathbf{K} + \sigma_n^2\mathbf{I})^{-1}\mathbf{K}_*,
    \label{eq:gp_var}
\end{align}
where $\mathbf{K} = k(\mathbf{t}, \mathbf{t})$,
$\mathbf{K}_* = k(\mathbf{t}, \mathbf{t}_*)$, and
$\mathbf{K}_{**} = k(\mathbf{t}_*, \mathbf{t}_*)$.

\subsubsection{Mat\'{e}rn Kernels}

The Mat\'{e}rn family of kernels provides control over the smoothness of
the GP sample paths via the parameter $\nu$:
\begin{equation}
  k_{\nu}(\Delta t) = \sigma_f^2 \frac{2^{1-\nu}}{\Gamma(\nu)}
    \left(\frac{\sqrt{2\nu}\,|\Delta t|}{\ell}\right)^\nu
    K_\nu\!\left(\frac{\sqrt{2\nu}\,|\Delta t|}{\ell}\right),
  \label{eq:matern}
\end{equation}
where $\ell$ is the length-scale, $\sigma_f^2$ the signal variance, and
$K_\nu$ the modified Bessel function. For $\nu = 5/2$, the kernel yields
twice-differentiable sample paths:
\begin{equation}
  k_{5/2}(\Delta t) = \sigma_f^2
    \left(1 + \frac{\sqrt{5}\,|\Delta t|}{\ell}
    + \frac{5\,\Delta t^2}{3\ell^2}\right)
    \exp\!\left(-\frac{\sqrt{5}\,|\Delta t|}{\ell}\right).
  \label{eq:matern52}
\end{equation}
This is our default temporal kernel for tumour volume dynamics, encoding the
prior belief that volume evolves smoothly over time.

\subsubsection{Connection to Linear Mixed-Effects Models}

A GP with a linear kernel $k(t, t') = \sigma_0^2 + \sigma_1^2 t t'$ and
zero mean is mathematically equivalent to a random-intercept, random-slope
LME \cite{rasmussen2006gaussian}. This connection is not merely formal:
it means that the LME baseline (Stage~1) is a special case of the GP
framework, and improvements from the GP must come from the nonlinear
component captured by the Mat\'{e}rn kernel. This nesting property ensures
that the GP can never perform worse than the LME on training data and
provides a principled model comparison via log-marginal likelihood.

\subsubsection{Hierarchical GP for Longitudinal Cohorts}

When multiple patients share similar growth dynamics, a hierarchical
(empirical Bayes) GP structure pools information across patients.
The Gaussian Process Panel Model (GPPM) framework of Karch et al.\
\cite{karch2020gppm} formalises this: shared kernel hyperparameters
$\boldsymbol{\theta} = (\sigma_f, \ell, \sigma_n)$ are estimated from the
pooled marginal likelihood across all patients, while patient-specific
posteriors arise from conditioning on individual observations. This is
particularly important at small $N$: per-patient kernel fitting with
$n_i \in [2, 6]$ observations is ill-conditioned, but pooling across
$N = 33$ patients with $\sim$100 total observations provides stable
hyperparameter estimates.

\subsubsection{Multi-Output GPs and the Intrinsic Coregionalisation Model}

Multi-Output Gaussian Processes (MOGPs) extend scalar GPs to vector-valued
functions $\mathbf{f}: \R \to \R^D$, capturing correlations across output
dimensions. The Intrinsic Coregionalisation Model (ICM) \cite{alvarez2012kernels}
parameterises the cross-output covariance as a Kronecker product:
\begin{equation}
  \mathbf{K}(\mathbf{t}, \mathbf{t}') =
    \mathbf{B} \otimes k_t(\mathbf{t}, \mathbf{t}'),
  \label{eq:icm}
\end{equation}
where $\mathbf{B} \in \R^{D \times D}$ is a positive semi-definite
coregionalisation matrix encoding inter-output correlations, and
$k_t$ is a scalar temporal kernel shared across outputs. The Linear Model of
Coregionalisation (LMC) generalises this to a sum over $Q$ latent functions:
\begin{equation}
  \mathbf{K}(\mathbf{t}, \mathbf{t}') =
    \sum_{q=1}^{Q} \mathbf{B}_q \otimes k_q(\mathbf{t}, \mathbf{t}').
  \label{eq:lmc}
\end{equation}

% ---------------------------------------------------------------------------
% 2.3 Nonlinear Mixed-Effects Models and Item Response Theory
% ---------------------------------------------------------------------------
\subsection{Nonlinear Mixed-Effects Models and Latent Severity}
\label{sec:bg:nlme}

\subsubsection{NLME Formulation}

Nonlinear mixed-effects (NLME) models extend the LME framework to nonlinear
growth functions while preserving the population--individual decomposition
\cite{pinheiro2000mixed}. For patient $i$ at timepoint $j$:
\begin{equation}
  y_{ij} = g(s_i, t_{ij}; \boldsymbol{\theta}) + \epsilon_{ij},
  \quad \epsilon_{ij} \sim \mathcal{N}(0, \sigma^2),
  \label{eq:nlme}
\end{equation}
where $g$ is a nonlinear growth function, $s_i$ is a patient-specific latent
variable (the ``severity'' or growth propensity), and $\boldsymbol{\theta}$
are population-level parameters.

\subsubsection{Reduced Gompertz Model}

Vaghi et al.\ \cite{vaghi2020reduced} demonstrated that for tumour growth
modelling, a \emph{reduced Gompertz} model with a single patient-specific
parameter explains the majority of inter-patient growth heterogeneity:
\begin{equation}
  V_i(t) = V_{i,0} \exp\!\left(\frac{\alpha_i}{\beta}
    \bigl(1 - e^{-\beta t}\bigr)\right),
  \label{eq:reduced_gompertz}
\end{equation}
where $\beta$ is a shared deceleration parameter and $\alpha_i$ is the
patient-specific initial growth rate. This aligns with Engelhardt et al.'s
finding \cite{engelhardt2023gompertz} that Gompertz dynamics best describe
meningioma growth. The single-parameter reduction is essential at small $N$:
with only $\sim$3.6 observations per patient on average, estimating more than
1--2 individual-level parameters per patient is infeasible.

\subsubsection{Connection to Item Response Theory}

The severity model has a formal connection to Item Response Theory (IRT)
\cite{embretson2000irt}. In the IRT analogy:
\begin{itemize}[leftmargin=*]
  \item Patients are ``examinees'' with latent ability $s_i \in [0, 1]$.
  \item Timepoints are ``items'' with difficulty determined by the elapsed
    time $t_{ij}$.
  \item Observed growth is the ``response'' governed by a monotonic link
    function.
\end{itemize}
Proust-Lima et al.\ \cite{proustlima2014joint,proustlima2023lcmm} developed
joint latent class mixed models (JLCM) that formalise this connection for
longitudinal biomarker trajectories. Our Stage~2 severity model can be viewed
as a single-class JLCM with a continuous (rather than discrete) latent
variable.

\subsubsection{Monotonic Constraints}

The severity variable $s_i$ must induce a monotonic relationship with
observed growth (higher severity implies faster growth at all times). This
can be enforced through:
\begin{itemize}[leftmargin=*]
  \item \textbf{Parametric constraints:} Using growth functions that are
    monotonically increasing in $s_i$ by construction (e.g., reduced
    Gompertz with $\alpha_i = f(s_i)$ where $f$ is monotone).
  \item \textbf{Constrained monotonic neural networks (CMNN):}
    Runje \& Shankaranarayana \cite{runje2023cmnn} provide architectures
    with provable monotonicity via non-negative weight constraints.
  \item \textbf{Monotonic GPs:} Riihim\"{a}ki \& Vehtari
    \cite{riihimaki2010monotonic} impose derivative constraints on the GP
    posterior to enforce monotonicity.
\end{itemize}

% ---------------------------------------------------------------------------
% 2.4 Sample Size Constraints
% ---------------------------------------------------------------------------
\subsection{Sample Size Constraints for Predictive Modelling}
\label{sec:bg:sample_size}

The longitudinal cohort in this work contains $N = 31$--$58$ patients with
approximately 112 total observations. This places severe constraints on model
complexity.

\subsubsection{Subjects-per-Variable Rules}

Riley et al.\ \cite{riley2019pmsampsize} provide sample size criteria for
prediction models, recommending a minimum of 15--20 subjects per predictor
parameter (events per variable, EPV) for stable coefficient estimation.
Harrell \cite{harrell2015regression} similarly advocates a maximum of
$N/15 \approx 2$--$3$ predictor parameters for $N = 31$--$58$. These rules
imply that the growth prediction model should have at most 2--3 free
parameters per patient, ruling out high-dimensional neural approaches.

\subsubsection{Effective Sample Size in Multilevel Models}

For hierarchical/mixed-effects models, the effective sample size depends on
both the number of patients $N$ and the number of observations per patient
$n_i$. Snijders \& Bosker \cite{snijders2012multilevel} note that
level-2 parameters (between-patient effects) depend on $N$, not on
$\sum n_i$. This is critical: even though we have $\sim$112 total
observations, the effective sample size for patient-level parameters remains
$N = 31$--$58$.

\subsubsection{Cross-Validation Error at Small $N$}

Varoquaux \cite{varoquaux2018cv} demonstrated that cross-validation error
estimates have standard deviations of approximately $\pm 10\%$ at $N = 100$.
At $N = 33$, the uncertainty is larger still. This motivates our use of LOPO-CV
(maximising training set size) and bootstrap confidence intervals, while
cautioning against over-interpreting small differences between models.

% ---------------------------------------------------------------------------
% 2.5 Foundation Models for Brain MRI
% ---------------------------------------------------------------------------
\subsection{Foundation Models for Brain MRI}
\label{sec:bg:foundation}

The paradigm of pretraining large encoder--decoder architectures on massive,
diverse datasets and subsequently fine-tuning on downstream tasks has
transformed medical image analysis \cite{he2023foundation,moor2023foundation}.
BrainSegFounder \cite{cox2024brainsegfounder} instantiates this paradigm for
3D brain MRI segmentation through a three-stage training pipeline:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Self-supervised pretraining.} A SwinUNETR encoder
    \cite{hatamizadeh2022swinunetr,tang2022self} is pretrained on 41,400+
    brain MRI subjects from the UK Biobank using a combination of masked image
    modelling, contrastive learning, and rotation prediction.

  \item \textbf{Supervised pretraining.} The encoder is paired with the
    SwinUNETR decoder and trained on the BraTS~2021 glioma segmentation
    benchmark \cite{baid2021rsna} at $128^3$ ROI.

  \item \textbf{Per-fold fine-tuning.} Final fine-tuning on BraTS~2021 with
    cross-validation yields the released checkpoints.
\end{enumerate}

\subsubsection{SwinUNETR Architecture}

SwinUNETR combines the Swin Transformer \cite{liu2021swin} with a
UNet-style encoder--decoder structure. The encoder consists of a patch
embedding layer followed by four transformer stages with shifted window
self-attention. For BSF-Tiny (the variant used in this work):

\begin{table}[H]
  \centering
  \caption{SwinUNETR encoder (BSF-Tiny) architecture. Input: $[B, 4, 128^3]$.}
  \label{tab:swinunetr_arch}
  \begin{tabular}{lcccc}
    \toprule
    \textbf{Stage} & \textbf{Output shape} & \textbf{Channels} & \textbf{Heads} & \textbf{Blocks} \\
    \midrule
    Patch embed & $[B, 48, 64^3]$  & 48  & --- & --- \\
    Stage 1     & $[B, 96, 32^3]$  & 96  & 3   & 2 \\
    Stage 2     & $[B, 192, 16^3]$ & 192 & 6   & 2 \\
    Stage 3     & $[B, 384, 8^3]$  & 384 & 12  & 2 \\
    Stage 4     & $[B, 768, 4^3]$  & 768 & 24  & 2 \\
    encoder10   & $[B, 768, 4^3]$  & 768 & --- & --- \\
    \bottomrule
  \end{tabular}
\end{table}

Global Average Pooling (GAP) over the spatial dimensions of \texttt{encoder10}
produces the feature vector $\mathbf{h} \in \R^{768}$ used in Stage~3 of our
pipeline.

% ---------------------------------------------------------------------------
% 2.6 Low-Rank Adaptation (LoRA)
% ---------------------------------------------------------------------------
\subsection{Low-Rank Adaptation}
\label{sec:bg:lora}

Low-Rank Adaptation (LoRA) \cite{hu2022lora} provides parameter-efficient
fine-tuning by injecting trainable rank-$r$ decompositions into frozen
pretrained weight matrices. For a pretrained weight matrix
$\mathbf{W}_0 \in \R^{d \times k}$, LoRA parameterises the adapted weight as:
\begin{equation}
  \mathbf{W} = \mathbf{W}_0 + \frac{\alpha}{r}\, \mathbf{B}\mathbf{A},
  \quad \mathbf{B} \in \R^{d \times r},\; \mathbf{A} \in \R^{r \times k},
  \label{eq:lora}
\end{equation}
where $r \ll \min(d, k)$ is the rank, $\alpha$ is a scaling hyperparameter,
$\mathbf{A}$ is initialised from $\mathcal{N}(0, \sigma^2)$, and
$\mathbf{B} = \mathbf{0}$ so that the adapted model initially reproduces the
pretrained model exactly.

\textbf{DoRA} \cite{liu2024dora} extends LoRA by decomposing the weight
update into magnitude and direction components:
\begin{equation}
  \mathbf{W} = m \cdot \frac{\mathbf{W}_0 + \mathbf{B}\mathbf{A}}
    {\|\mathbf{W}_0 + \mathbf{B}\mathbf{A}\|_c},
  \label{eq:dora}
\end{equation}
where $m$ is a learnable magnitude vector and $\|\cdot\|_c$ denotes
column-wise normalisation. We evaluate DoRA as an ablation condition
(A5, \cref{sec:eval:ablations}).

In the context of our three-stage ladder, LoRA adaptation is relevant to
Stage~3 (representation learning). It is also relevant to Stage~1 insofar
as an adapted encoder may produce better segmentations and hence more
accurate volume estimates. We evaluate this explicitly in ablation~A1
(\cref{sec:eval:ablations}).

% ---------------------------------------------------------------------------
% 2.7 Dimensionality Reduction for Deep Features
% ---------------------------------------------------------------------------
\subsection{Dimensionality Reduction for Deep Features}
\label{sec:bg:dimred}

The BrainSegFounder encoder produces 768-dimensional feature vectors, which
must be compressed for use in growth prediction models at small $N$.

\subsubsection{PCA with Supervised Selection}

Principal Component Analysis provides an unsupervised linear projection to
a lower-dimensional subspace. For Stage~3, we apply PCA within each LOPO-CV
fold (to prevent information leakage) and select 3--6 components, guided
by: (i)~the elbow of the variance-explained curve; (ii)~correlation with
the volume growth target; and (iii)~the sample size constraint of 2--3
parameters per predictor.

\subsubsection{ARD Kernels}

Automatic Relevance Determination (ARD) kernels assign a separate length-scale
$\ell_d$ per input dimension \cite{rasmussen2006gaussian}. During marginal
likelihood optimisation, irrelevant dimensions acquire large length-scales
(effectively deactivating them). ARD provides an alternative to explicit
dimensionality reduction, performing implicit feature selection within the GP.
However, at $N = 33$ the number of ARD hyperparameters must be carefully
controlled to avoid overfitting.

\subsubsection{Active Subspaces}

Tripathy et al.\ \cite{tripathy2016active} formalise the notion of active
subspaces for dimensionality reduction in GP regression, identifying the
low-dimensional linear subspace of the input that captures most of the
variation in the function output. This provides a principled alternative to
PCA that accounts for the target variable, and is conceptually related to the
SDP's supervised projection in Stage~3.

% ---------------------------------------------------------------------------
% 2.8 Disentangled Representation Learning
% ---------------------------------------------------------------------------
\subsection{Disentangled Representation Learning}
\label{sec:bg:disentanglement}

The impossibility theorem of Locatello et al.\ \cite{locatello2019challenging}
established that unsupervised disentanglement is theoretically impossible
without inductive biases. Their follow-up work
\cite{locatello2020disentangling} demonstrated that as few as 100 labelled
examples suffice to resolve this impossibility through supervision.

These results motivate our Supervised Disentangled Projection (SDP) approach
in Stage~3: rather than relying on unsupervised disentanglement, we use
explicit semantic supervision from segmentation-derived features to anchor
each latent partition to a known generative factor. The SDP combines semantic
regression, VICReg-style \cite{bardes2022vicreg} variance--covariance
regularisation, and distance correlation \cite{szekely2007measuring}
penalties. Spectral normalisation \cite{miyato2018spectral} on all linear
layers bounds the global Lipschitz constant, ensuring that encoder feature
noise is not amplified through the projection
\cite{gouk2021regularisation}. Full details are in \cref{sec:sdp}.

% ---------------------------------------------------------------------------
% 2.9 Related Work
% ---------------------------------------------------------------------------
\subsection{Related Work}
\label{sec:bg:related}

\subsubsection{Brain Tumour Growth Prediction}

Tumour growth prediction from longitudinal MRI has been addressed through
several paradigms. Classical approaches fit parametric growth models
(exponential, Gompertz, logistic) to volume trajectories
\cite{nakasu2005serial,hashimoto2012natural}. Reaction--diffusion models
\cite{swanson2000quantitative,hormuth2021image} incorporate spatial diffusion
but require tissue-specific parameterisation and are computationally expensive.

Recent deep learning approaches include spatio-temporal convolutional LSTMs
\cite{zhang2020spatio} achieving 83.2\% Dice on glioma growth prediction,
GP-GAN \cite{elazab2020gp} combining Gaussian processes with GANs, and neural
fields conditioned on time \cite{chen2024vestibular} for vestibular schwannoma
growth. Puglisi et al.\ \cite{puglisi2025brain} demonstrated that a linear
mixed-effects model in a learned latent space produces clinically meaningful
progression predictions for brain atrophy, trained on 11,730~MRIs.

For meningiomas specifically, most existing work is limited to retrospective
volumetric analysis \cite{hashimoto2012natural,ius2020management,
nakamura2003natural}. To our knowledge, no prior work has provided
out-of-sample prediction benchmarks for meningioma growth.

\subsubsection{Latent Space Temporal Prediction}

The Joint Embedding Predictive Architecture (JEPA) framework
\cite{lecun2022path} advocates predicting in representation space rather than
pixel space. Puglisi et al.\ \cite{puglisi2025brain} validated this principle
for brain atrophy using LME in a learned latent space. Our Stage~3 aligns
with this ``predict in latent space'' principle while our Stages~1 and~2 serve
as necessary baselines to determine whether the latent space provides value
beyond simpler volumetric features.

\subsubsection{Domain Adaptation in Medical Imaging}

Domain adaptation for medical image analysis has been addressed through
adversarial training \cite{kamnitsas2017unsupervised}, style transfer
\cite{chen2022maxstyle}, and parameter-efficient fine-tuning of foundation
models \cite{dutt2023parameter}. Ben-David et al.\ \cite{bendavid2010theory}
established the theoretical framework relating domain divergence to target
risk. Our LoRA adaptation (Stage~3) contributes empirical evidence on the
marginal value of domain adaptation when a strong SSL-pretrained encoder is
available.
