% =============================================================================
% Section 10: Conclusion
% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

% ---------------------------------------------------------------------------
% 10.1 Summary of Contributions
% ---------------------------------------------------------------------------
\subsection{Summary of Contributions}
\label{sec:conc:summary}

This thesis has presented a three-stage complexity ladder for meningioma
growth forecasting from longitudinal multi-modal 3D MRI. Rather than
proposing a single complex pipeline and reporting its metrics, we
systematically evaluate whether each successive level of model
sophistication provides measurable out-of-sample improvement at $N = 30$
patients. The key contributions and findings are:

\begin{enumerate}[leftmargin=*]
  \item \textbf{The first out-of-sample meningioma growth prediction
    benchmark.} Using LOPO-CV on a longitudinal cohort of 30~patients,
    we provide the first reported $R^2$, MAE, and calibration metrics for
    meningioma volume prediction. The best model---a Linear Mixed-Effects
    model on volumes from a decoder-adapted BrainSegFounder
    segmentation---achieves $R^2 = 0.387$ with 90\% calibration coverage
    under the last-from-rest prediction protocol.

  \item \textbf{Segmentation source dominates model complexity.}
    A systematic comparison across four segmentation sources (manual,
    frozen BrainSegFounder, decoder-adapted, LoRA-adapted) reveals that
    segmentation quality is a stronger determinant of growth prediction
    than model class. The same LME model yields $R^2 = 0.028$ on manual
    segmentations versus $R^2 = 0.387$ on adapted segmentations---a
    14-fold improvement from the data pipeline alone.

  \item \textbf{LoRA adaptation: marginal for features, catastrophic for
    segmentation.} While LoRA achieves only $\Delta R^2 = +0.016$ for
    downstream feature quality on semantic probes, the LoRA-adapted
    segmentation is catastrophically broken (WT Dice $= 0.003$),
    producing degenerate growth predictions ($R^2 = -94$). This finding
    highlights the importance of evaluating adapters in the context of
    their downstream task, not just proxy metrics.

  \item \textbf{Stage~2 severity model fails its gate---ordinal time is
    the blocker.} The latent severity model achieves $R^2 = -3.54$,
    substantially worse than Stage~1. Root cause analysis identifies the
    ordinal time representation as the primary failure mode: without exact
    timestamps, the quantile transform cannot distinguish patients with
    different growth rates. This is a \emph{finding} that demonstrates
    the complexity ladder's gate mechanism working as designed---preventing
    a failing model from propagating errors.

  \item \textbf{The complexity ladder as a methodological framework.}
    The staged approach with explicit gate criteria ($\Delta R^2 > 0$,
    $p < 0.05$, CI excluding zero) provides a principled answer to the
    question ``how much complexity is warranted?'' At $N = 30$, the answer
    is: a simple LME on well-segmented volumes. The framework is
    transferable to other small-sample longitudinal prediction problems.
\end{enumerate}

% ---------------------------------------------------------------------------
% 10.2 Limitations
% ---------------------------------------------------------------------------
\subsection{Limitations}
\label{sec:conc:limitations}

\subsubsection{Ordinal Time Is the Critical Blocker}

The absence of exact acquisition timestamps is the single most impactful
limitation. Ordinal time indices ($1, 2, 3, \ldots$) destroy temporal
magnitude information, preventing the severity model from learning meaningful
growth dynamics and rendering temporal hyperparameters (GP length-scales,
growth rates) clinically uninterpretable. This limitation does not invalidate
the Stage~1 comparison (all models share the same time representation) but
blocks all Stage~2+ analyses.

\subsubsection{Cohort Size}

The longitudinal cohort ($N = 30$ patients, $\sim$112~observations) is a
fundamental bottleneck. With approximately 57.6\% of patients having only
2~studies, per-patient growth trajectories are sparse. The GP and LME
frameworks handle this via automatic shrinkage toward the population mean,
but predictions for 2-observation patients are necessarily dominated by
population-level dynamics rather than individual-specific growth rates.
The sample size also limits the statistical power of the variance
decomposition: Varoquaux \cite{varoquaux2018cv} showed that CV error
estimates have standard deviations of $\sim$$\pm$10\% at $N = 100$, and
uncertainty is larger still at $N = 30$.

\subsubsection{No Bootstrap Confidence Intervals}

The current results do not include bootstrap CIs on the reported $R^2$
values. The $R^2 = 0.387$ of the best model may have a CI that includes
zero, which would weaken the claim of genuine predictive ability.
Computing BCa bootstrap CIs is a priority for the final evaluation.

\subsubsection{LoRA Segmentation Failure}

The catastrophic failure of LoRA-adapted segmentation (WT Dice $= 0.003$)
is unexpected given that the same LoRA checkpoint achieves reasonable
segmentation Dice on BraTS-MEN validation data. This suggests a mismatch
between the LoRA training data distribution and the longitudinal cohort,
possibly due to preprocessing differences or domain shift between the
BraTS-MEN competition data and the clinical Andalusian cohort.

\subsubsection{Single-Site Longitudinal Data}

The longitudinal cohort is collected from hospitals in a single region
(Andalusia, Spain), which may limit generalisability to other populations,
scanner manufacturers, and acquisition protocols. ComBat harmonisation
addresses scanner batch effects but cannot correct for biological population
differences.

\subsubsection{Absence of Clinical Validation}

The pipeline has not been validated against clinical outcomes (intervention
decisions, actual tumour behaviour at longer follow-up). Volume prediction
$R^2$ and calibration metrics are necessary but not sufficient for clinical
utility.

% ---------------------------------------------------------------------------
% 10.3 Future Work
% ---------------------------------------------------------------------------
\subsection{Future Work}
\label{sec:conc:future}

\subsubsection{Exact Timestamps (Highest Priority)}

Obtaining exact acquisition dates would: (i)~enable re-evaluation of Stage~2,
which currently fails due to ordinal time; (ii)~provide physically meaningful
growth rate estimation; (iii)~improve GP length-scale interpretability; and
(iv)~allow direct comparison with volumetric growth rates reported in the
clinical literature \cite{hashimoto2012natural,nakamura2003natural,
engelhardt2023gompertz}.

\subsubsection{Bootstrap Confidence Intervals}

Computing BCa bootstrap CIs for all reported $R^2$ values will quantify the
uncertainty of our findings and determine whether the $R^2 = 0.387$ of the
best model is statistically distinguishable from zero. This is essential for
the variance decomposition analysis.

\subsubsection{LoRA Segmentation Diagnosis}

Investigating the catastrophic LoRA segmentation failure on the longitudinal
cohort---despite acceptable performance on BraTS-MEN validation---would
clarify whether this is a data preprocessing issue, a domain shift problem,
or a fundamental limitation of LoRA adaptation for cross-cohort transfer.

\subsubsection{Cohort Expansion to $N \geq 50$}

Multi-centre collaborations targeting $N \geq 50$ patients would: (i)~tighten
the variance decomposition CIs, enabling confident $\Delta R^2$ conclusions;
(ii)~potentially unlock value in Stage~2 and~3 by providing sufficient data
for more complex models; (iii)~enable subgroup analyses by tumour grade,
location, and patient demographics.

\subsubsection{Clinical Decision Support Integration}

The GP posterior's calibrated uncertainty is directly suitable for clinical
decision support: the probability that a tumour will exceed a volume
threshold $V_{\text{critical}}$ within a specified time horizon can be
computed analytically from the Gaussian predictive distribution:
\begin{equation}
  P(V(t_{\text{future}}) > V_{\text{critical}}) =
    1 - \Phi\!\left(\frac{\log(V_{\text{critical}} + 1) - \mu_*(t_{\text{future}})}
    {\sigma_*(t_{\text{future}})}\right),
  \label{eq:clinical_decision}
\end{equation}
where $\Phi$ is the standard normal CDF and $(\mu_*, \sigma_*)$ are the GP
posterior mean and standard deviation in log-volume space. Integrating this
into a clinical workflow would require prospective validation and regulatory
approval.

\subsubsection{Foundation Model Scaling}

BrainSegFounder is a relatively small foundation model (BSF-Tiny, 62M
parameters). Larger variants or more recent foundation models pretrained on
broader datasets may provide stronger initial features, potentially shifting
the Stage~3 $\Delta R^2$ from marginal to significant. However, this
hypothesis must be tested empirically rather than assumed.
