% =============================================================================
% Section 1: Introduction
% =============================================================================
\section{Introduction}
\label{sec:introduction}

% ---------------------------------------------------------------------------
% 1.1 Clinical Motivation
% ---------------------------------------------------------------------------
\subsection{Clinical Motivation}
\label{sec:intro:clinical}

Meningiomas are the most common primary intracranial tumours, accounting for
approximately 39\% of all central nervous system neoplasms, with an
age-adjusted incidence rate of 9.51 per 100,000 population
\cite{ostrom2023brain}. The majority are classified as WHO Grade~I (benign),
and a substantial fraction are discovered incidentally on neuroimaging
performed for unrelated indications \cite{vernooij2007incidental}. For these
incidentally discovered meningiomas, clinical management hinges on a
deceptively simple question: \emph{will this tumour grow?}

Current clinical practice relies on serial MRI surveillance with manual
volumetric assessment at intervals of 3--12 months
\cite{hashimoto2012natural,ius2020management}. This approach has three
fundamental limitations. First, manual volumetry is labour-intensive and
subject to considerable inter-observer variability, with reported coefficient
of variation exceeding 15\% for small tumours \cite{defined2016interobserver}.
Second, the decision to intervene (surgery or radiotherapy) is typically made
reactively---after significant growth has already occurred---rather than
proactively based on predicted trajectories. Third, meningioma growth dynamics
are heterogeneous: approximately 67\% show measurable growth at 4--5~year
follow-up \cite{hashimoto2012natural}, but growth patterns range from stable
quiescence through steady linear expansion to Gompertz-type saturation
\cite{engelhardt2023gompertz,fountain2017growth}, and no established imaging
biomarker reliably predicts which pattern a given tumour will follow.
Volumetric assessment substantially outperforms diameter-based measurements for
detecting growth, with 15.2\% sensitivity versus 5.6\% using linear diameter
\cite{benjamin2021volumetric}.

Quantitative growth prediction from longitudinal MRI could transform
meningioma management by enabling personalised surveillance schedules,
early identification of aggressive growth patterns, and evidence-based
timing of intervention. However, building such predictive models faces a
data bottleneck: longitudinal meningioma MRI datasets are rare, typically
small (tens of patients), and collected retrospectively across heterogeneous
scanners and protocols.

% ---------------------------------------------------------------------------
% 1.2 The Core Question
% ---------------------------------------------------------------------------
\subsection{The Core Question: Complexity versus Parsimony at Small~$N$}
\label{sec:intro:question}

The central question of this thesis is not merely ``can we predict meningioma
growth?'' but rather: \emph{at what level of model complexity does additional
sophistication stop improving out-of-sample prediction when the longitudinal
cohort contains only $N = 31$--$58$ patients?}

This question is both practical and methodological. On the practical side,
the answer determines whether clinical growth prediction requires expensive
deep learning pipelines or whether simpler segmentation-based volumetric
models suffice. On the methodological side, it addresses a widespread gap in
the medical imaging literature, where increasingly complex architectures are
proposed without systematic comparison against simpler baselines under the
same evaluation protocol.

A key constraint is the effective sample size. With $N = 31$--$58$ patients
averaging approximately 3.6 observations each ($\sim$112 total observations),
the maximum number of predictor parameters that can be reliably estimated is
2--3 at the most conservative rule of thumb of 15--20 subjects per predictor
\cite{riley2019pmsampsize,harrell2015regression}. This places a hard ceiling
on model complexity: any model that exceeds this budget must earn its place
through demonstrably superior out-of-sample performance.

To our knowledge, no published work provides out-of-sample prediction
benchmarks for meningioma growth. Existing studies are limited to
retrospective volumetric characterisation and within-sample growth curve
fitting \cite{hashimoto2012natural,nakamura2003natural,engelhardt2023gompertz}.
This thesis addresses that gap directly.

% ---------------------------------------------------------------------------
% 1.3 Experimental Design: A Three-Stage Complexity Ladder
% ---------------------------------------------------------------------------
\subsection{Experimental Design: A Three-Stage Complexity Ladder}
\label{sec:intro:approach}

Rather than proposing a single pipeline and reporting its metrics, we organise
the experimental programme as a \emph{three-stage complexity ladder} in which
each stage must justify its existence through measurable improvement over the
preceding one. A cross-cutting variance decomposition quantifies the marginal
$\Delta R^2$ contributed by each stage.

\begin{description}[leftmargin=*,labelwidth=!]
  \item[Stage 1 --- Segmentation-Based Volumetric Baseline (\cref{sec:growth:stage1}).]
    The simplest credible approach: segment the tumour from each MRI
    (using a frozen or adapted encoder), extract whole-tumour volume in mm$^3$,
    and model the resulting scalar volume trajectories with GP and LME models.
    This stage requires only a reliable segmentation model and a single scalar
    per observation. Three models of increasing flexibility are evaluated:
    a pooled scalar GP (ScalarGP), a linear mixed-effects model (LME), and a
    hierarchical GP with population mean from the LME (HGP). An optional
    Gompertz mean function is assessed as an alternative to the linear
    population trend.

  \item[Stage 2 --- Latent Severity Model (\cref{sec:growth:stage2}).]
    A nonlinear mixed-effects (NLME) model posits a single latent severity
    variable $s_i \in [0, 1]$ per patient, estimated via quantile
    transformation of growth patterns. This is formally connected to Item
    Response Theory (IRT): patients are ``examinees'' with latent ability
    $s_i$, timepoints are ``items'' with difficulty determined by $t_{ij}$,
    and observed growth is the ``response'' governed by a monotonic link
    function \cite{vaghi2020reduced,proustlima2014joint}. Stage~2 adds at
    most 1--2 parameters per patient beyond Stage~1 while providing a richer
    individual-level characterisation.

  \item[Stage 3 --- Representation Learning (\cref{sec:growth:stage3}).]
    The full deep learning pipeline: BrainSegFounder
    \cite{cox2024brainsegfounder} encoder adapted via LoRA
    \cite{hu2022lora} (\cref{sec:lora}), followed by a Supervised
    Disentangled Projection \cite{locatello2019challenging} (\cref{sec:sdp})
    mapping $\mathbf{h} \in \R^{768}$ to $\mathbf{z} \in \R^{128}$, with
    PCA-compressed latent features fed to GP models with Automatic Relevance
    Determination (ARD) kernels. This is the most complex stage and the last
    to be attempted: it must demonstrate that high-dimensional deep features
    capture growth-relevant information beyond what volume and severity
    already explain.
\end{description}

The critical design principle is that \textbf{each stage must earn its place}.
Stage~2 is only warranted if it improves upon Stage~1. Stage~3 is only
warranted if it improves upon Stage~2. If volume alone captures the dominant
growth signal---as parsimony would suggest at $N = 31$--$58$---then the
primary scientific contribution is establishing that fact rigorously, not
building a more complex model.

% ---------------------------------------------------------------------------
% 1.4 Contributions
% ---------------------------------------------------------------------------
\subsection{Contributions}
\label{sec:intro:contributions}

This thesis makes the following contributions:

\begin{enumerate}[leftmargin=*]
  \item \textbf{The first out-of-sample meningioma growth prediction
    benchmark.} Using Leave-One-Patient-Out Cross-Validation (LOPO-CV) on a
    longitudinal cohort of 31--58 patients, we provide the first reported
    $R^2$, MAE, and calibration metrics for meningioma volume prediction.
    The volumetric baseline (Stage~1) establishes a reference against which
    all subsequent complexity can be measured.

  \item \textbf{A latent severity framework for growth characterisation.}
    We formulate meningioma growth heterogeneity as a latent variable problem
    connected to IRT and reduced Gompertz dynamics
    \cite{vaghi2020reduced}, where a single patient-specific severity
    parameter $s_i$ explains inter-patient variability. This provides
    interpretable patient stratification while adding minimal parameters.

  \item \textbf{A systematic assessment of representation learning value.}
    By placing the full BrainSegFounder + LoRA + SDP pipeline as Stage~3
    of the complexity ladder, we directly quantify the marginal $\Delta R^2$
    of deep features over volumetric and severity baselines. This addresses
    the recurring question in medical imaging of whether foundation model
    features improve upon simpler alternatives for specific downstream tasks.

  \item \textbf{Variance decomposition as an analytical framework.}
    The cross-cutting analysis decomposes prediction variance across the
    three stages, providing not only a model selection criterion but also a
    methodological template for evaluating complexity--benefit tradeoffs in
    small-sample longitudinal prediction.
\end{enumerate}

% ---------------------------------------------------------------------------
% 1.5 Document Structure
% ---------------------------------------------------------------------------
\subsection{Document Structure}

The remainder of this report is organised as follows.
\Cref{sec:background} reviews the theoretical foundations: meningioma growth
biology, GP regression for longitudinal data, nonlinear mixed-effects models,
sample size constraints, and representation learning.
\Cref{sec:data} describes the datasets, preprocessing, and data splits.
\Cref{sec:domain_gap} presents the domain gap analysis motivating LoRA
adaptation.
\Cref{sec:lora} details the LoRA encoder adaptation (Phase~1, relevant to
Stage~3).
\Cref{sec:sdp} describes the Supervised Disentangled Projection (Phase~2,
relevant to Stage~3).
\Cref{sec:encoding} covers cohort encoding and ComBat harmonisation
(Phase~3).
\Cref{sec:growth} presents the three-stage complexity ladder for growth
prediction, which is the core methodological contribution.
\Cref{sec:evaluation} consolidates the variance decomposition, quality
targets, and ablation results.
\Cref{sec:conclusion} discusses limitations and future directions.
